---
title: "TidyVerse CREATE"
author: "Kevin Havis"
date: "2024-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidy operations across columns

In this vignette we will walk through a useful function in the `dplyr` package for column-wise operations; `across()`.

```{r}
library(tidyverse)

# Load initial data

df <- as_tibble(trees)
```

`across()` is best paired with `mutate()`. Consider we would like to rank the respective `Girth`, `Height`, and `Volume` of the black cherry trees in the `trees` data.

```{r}
# Without across()
df |> 
  mutate(
    girth_rank = rank(Girth),
    height_rank = rank(Height),
    vol_rank = rank(Volume)
  )
```

This is easy enough but what if we had many columns? Or we did not know how many columns we are going to operate across?

We can use `across()` in conjunction with `mutate()` to perform this operation.

```{r}
# With across()
df |> 
  mutate(
    across(
      .cols = everything(),    # Operate across all columns
      .fns = rank,              # The function(s) to perform
      .names = "{.col}_rank"   # How the new columns should be named
    )
  )
```

We can also specify what *type* of columns we would like to perform by replacing `everything()` with `starts_with()`. You could also use `is.character` or `is.numerical`, which is useful for things such as string normalization, rounding floats into integers, etc.

```{r}
# Using across for specific columns
df |> 
  mutate(
    across(
      .cols = starts_with(c("Height", "Volume")),    # Operate on subset of cols
      .fns = rank,                       
      .names = "{.col}_rank"            
    )
  )
```
```{r}
# Operate on specific column type
df |> 
  mutate(
    tree = "I'm not a tree..."
  ) |> 
  mutate(
    across(
      .cols = where(is.character),  # Select character columns only
      .fns = ~ "I'm a tree!"               # Replace values with "tree"
    )
  )
```

```{r}
# Round numeric columns to integers

df |> 
  mutate(
    across(
      .cols = where(is.numeric),  # Select all numeric columns
      .fns = ~ round(.x),         # Apply round() to each column
      .names = "{.col}_int"       # Create new columns with _int suffix
    )
  )
```

