---
title: "Week3 Extra Credit"
author: "John Ferrara"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Extra Credit: Global Baseline Estimates

Using the information you collected on movie ratings, implement a Global Baseline Estimate recommendation system in R.   The attached spreadsheet provides the implementation algorithm.

Most recommender systems use personalized algorithms like “content management” and “item-item collaborative filtering.” Sometimes non-personalized recommenders are also useful or necessary. One of the best non-personalized recommender system algorithms is the “Global Baseline Estimate.

The job here is to use the survey data collected and write the R code that makes a movie recommendation using the Global Baseline Estimate algorithm. 

## Connection to the SQL Server 

Connecting the the localhost SQL server in order to pull in my movies data. 

#### Step 1 - Importing and installing the packages needed
```{r sql_connect1}

#install.packages("RMySQL")
library(RMySQL)
library(readr)
library(dplyr)
library(tidyr)

```


#### Step 2 - Establish connection to MySQL database
```{r sql_connect2}

# Read the password in from file
pw_path <- "~/Documents/Github/CUNYSPS_DATA607/secrets/mysql_root_pw.txt"
password <- read_file(path.expand(pw_path))

con <- dbConnect(RMySQL::MySQL(),
                 dbname = "assignment2_movies",
                 host = "127.0.0.1",
                 port = 3306,
                 user = "root",
                 password = password)
```

#### Step 3 - Listing tables in the selected schema to confirm the connection
```{r sql_connect3}
dbListTables(con)
```

## Pulling in Data 
Now that the connection is established, pulling in the data to a dataframe. 

```{r dataframe}
movies_df <- dbReadTable(con, "critics_ratings")

head(movies_df)
```

## Step 4 - Establish the Problem
With all of the data pulled into r, the next step is the establish the issue we're attempting to resolve. The problem is we want to get a proper movie rating  estimate for Lisa for for the movie Arrival.

## Step 5 - Get the  Movie Average
``` {r MovieAverages}
#Getting The avg for each movie
GoodfellasAVG <- mean(movies_df$Goodfellas, na.rm=TRUE)
VforVendettaAVG<-mean(movies_df$VforVendetta, na.rm=TRUE)
ArrivalAVG<-mean(movies_df$Arrival, na.rm=TRUE)
InceptionAVG<-mean(movies_df$Inception, na.rm=TRUE)
TheWitchAVG<- mean(movies_df$TheWitch, na.rm=TRUE)
TheDarkKnightAVG <- mean(movies_df$TheDarkKnight, na.rm=TRUE)
```

## Step 6 - Get the Mean for all of Users' Movie Ratings
```{r lisa_ratingavg}
## Getting the avg for each person
movies_df$mean_rating <- rowMeans(movies_df[, -1], na.rm = TRUE)
movies_df 

## Getting the overall avg of every rating
overall_mean_df <- movies_df %>%
  pivot_longer(cols = Goodfellas:TheDarkKnight, values_drop_na = TRUE) %>%
  summarize(mean_rating = mean(value, na.rm = TRUE))
overall_mean_df
# Limiting to value
overall_mean<- overall_mean_df$mean_rating
overall_mean
```

## Step 5 - Get the Global Baseline Estimate Formula  and then Derive what Lisa would rate Arrival

Following the formula in provided Excel file for Extra Credit Assignment. 

```{r solution}
#Getting the avg for lisa value 
lisa_avg_df <- movies_df%>%
  filter(critic == 'Lisa')
lisa_avg_df

lisa_avg = lisa_avg_df$mean_rating
lisa_avg

#Getting the differences
lisa_diferential <- lisa_avg-overall_mean
movie_diferential <-ArrivalAVG - overall_mean
 
## This is the estimat rating for Lisa for Arrival
lisa_est_rating_arrival <- overall_mean - abs(movie_diferential) - abs(lisa_diferential)
lisa_est_rating_arrival
```