---
title: "DATA607_Week9_Assignment"
author: "John Ferrara"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Web APIs

The New York Times web site provides a rich set of APIs, as described here: https://developer.nytimes.com/apis
You’ll need to start by signing up for an API key.
Your task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and
transform it into an R DataFrame.

```{r hhtr}
# install.packages("httr2")
library(httr2)
library(rjson)
library(readr)
library(tibble)
library(dplyr)

```

## Prepping Creds and oter setup
```{r set_up}
api_path <- "~/OneDrive/Desktop/CUNY_SPS_Materials/DATA607 - DATA ACQ/Week9/nyt_api_creds.json"
api_file <- read_file(path.expand(api_path))
json_data <- fromJSON(api_file)

```

```{r api Call}

### Gettign the most viewed article in the last 30 days. 
day_period <- "7"

url_call <- sprintf("https://api.nytimes.com/svc/mostpopular/v2/viewed/%s.json", day_period)

response <- request(url_call) %>%
  req_url_query("api-key" = json_data$public_key) %>%  # Add the API key as a query parameter
  req_perform()

# Parse the response content
data <- response %>% resp_body_json()

```


``` {r processing to df}

df_list <- list()

for (i in seq_along(data$results)) {
  article <- data$results[[i]]
  ## Taking Article informstion
  new_row <- data.frame(
    uri = article$uri,
    url = article$url,
    id = article$id,
    asset_id = article$asset_id,
    source = article$source,
    published_date = article$published_date,
    updated = article$updated,
    section = article$section,
    subsection = article$subsection,
    ## THis value can be null in my results and im accomodating
    nytsection = ifelse(!is.null(article$nytsection), article$nytsection, NA),
    adx_keywords = article$adx_keywords,
    byline = article$byline,
    type = article$type,
    title = article$title,
    abstract = article$abstract,
    stringsAsFactors = FALSE
  )
  
  #Append the new row to the list
  df_list[[i]] <- new_row
  
}

df <- bind_rows(df_list)
df

```


```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
