---
title: "DATA607_Week10_Assgnmt"
author: "John Ferrara"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Week 10 Assignment

#### Instructions
In Text Mining with R, Chapter 2 looks at Sentiment Analysis.  In this assignment, you should start by getting the primary example code from chapter 2 working in an R Markdown document.  You should provide a citation to this base code.  You’re then asked to extend the code in two ways:

Work with a different corpus of your choosing, and
Incorporate at least one additional sentiment lexicon (possibly from another R package that you’ve found through research).
As usual, please submit links to both an .Rmd file posted in your GitHub repository and to your code on rpubs.com.  You make work on a small team on this assignment.

## Base Code from Text Book

```{r part1}

## Base Code Citation: https://www.tidytextmining.com/sentiment.html


# install.packages("janeaustenr")
#install.packages("tidytext")

library(tidytext)
library(janeaustenr)
library(dplyr)
library(stringr)

tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, 
                                regex("^chapter [\\divxlc]", 
                                      ignore_case = TRUE)))) %>%
  ungroup() %>%
  unnest_tokens(word, text)

head(tidy_books)

```
### Sentiment Analysis on Another Data Set found
```{r step2}
#install.packages("textdata")

library(textdata)
# IMBD Reviews
imdb_data <- textdata::dataset_imdb()
head(imdb_data)

print("test")
```




