---
title: "Assignment2 - SQL Movies Data"
author: "John Ferrara"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task: Pull In a Custom SQL Data Table into An R Dataframe

This is the second file of the two files that is part if this assignment. The first file is a SQL file (Assignment2.sql) that creates the custom schema and table for use in R Studio.

## Connection to the SQL Server 

Connecting the the localhost SQL server in order to pull in my movies data. 

#### Step 1 - Importing and installing the packages needed
```{r sql_connect1}

#install.packages("RMySQL")
library(RMySQL)
library(readr)

```

#### Step 2 - Establish connection to MySQL database
```{r sql_connect2}

# Read the password in from file
pw_path <- "~/Documents/Github/CUNYSPS_DATA607/secrets/mysql_root_pw.txt"
password <- read_file(path.expand(pw_path))

con <- dbConnect(RMySQL::MySQL(),
                 dbname = "assignment2_movies",
                 host = "127.0.0.1",
                 port = 3306,
                 user = "root",
                 password = password)
```

#### Step 3 - Listing tables in the selected schema to confirm the connection
```{r sql_connect3}
dbListTables(con)
```

## Pulling in Data 
Now that the connection is established, pulling in the data to a dataframe. 

```{r dataframe}
movies_df <- dbReadTable(con, "movies")

head(movies_df)
```

## Demonstrating working with Missing Data
For the purposes of this assignment, the last movie provided does not have a rating. The Dark Knight has no rating, and the value in that cell is null. If one were to get an average of the the ratings for the movies listed in this database, the Null value would need to be ignored. 

```{r average_rating}

# Doing it this way yields a null because of the single null value
avg_rating <- mean(movies_df$rating)

#Doing it this way removed the null value from the calculation
avg_rating <- mean(movies_df$rating, na.rm = TRUE)

print(avg_rating)

```
