---
title: "DATA624_Homework6"
author: "John Ferrara"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(fpp3)
```

## 1) Figure 9.32 shows the ACFs for 36 random numbers, 360 random numbers and 1,000 random numbers.

[Figure 9.32]("Figure_932.png")





### a) Explain the differences among these figures. Do they all indicate that the data are white noise?

```{r Exercise1a}

```
### b) Why are the critical values at different distances from the mean of zero? Why are the autocorrelations different in each figure when they each refer to white noise?

```{r Exercise1b}

```

## 2) A classic example of a non-stationary series are stock prices. Plot the daily closing prices for Amazon stock (contained in gafa_stock), along with the ACF and PACF. Explain how each plot shows that the series is non-stationary and should be differenced.

``` {r Exercise2}

```

## 3) For the following series, find an appropriate Box-Cox transformation and order of differencing in order to obtain stationary data.

### a) Turkish GDP from global_economy.
```{r Exercise3a}

```
### b) Accommodation takings in the state of Tasmania from aus_accommodation.
```{r Exercise3b}

```
### c) Monthly sales from souvenirs.
```{r Exercise3c}

```

## 5) For your retail data (from Exercise 7 in Section 2.10), find the appropriate order of differencing (after transformation if necessary) to obtain stationary data.
``` {r exercise5}
```

## 6) Simulate and plot some data from simple ARIMA models.

### a)Use the following R code to generate data from an AR(1) model with ϕ1 = 0.6 and  σ2=1. The process starts with  y1=0.

``` {r exercise6a_reference}
#y <- numeric(100)
#e <- rnorm(100)
#for(i in 2:100)
#  y[i] <- 0.6*y[i-1] + e[i]
#sim <- tsibble(idx = seq_len(100), y = y, index = idx)
```

``` {r exercise6a}
```

### b) Produce a time plot for the series. How does the plot change as you change ϕ1 ?
``` {r exercise6b}
```

### c)Write your own code to generate data from an MA(1) model with θ1=0.6 and σ2=1.
``` {r exercise6c}
```

### d)Produce a time plot for the series. How does the plot change as you change θ1?
``` {r exercise6d}
```

### e) Generate data from an ARMA(1,1) model with ϕ1=0.6, θ1=0.6 and σ2=1.
``` {r exercise6e}
```

### f) Generate data from an AR(2) model with ϕ1=−0.8, ϕ2=0.3 and σ2=1. (Note that these parameters will give a non-stationary series.)
``` {r exercise6f}
```

### g) Graph the latter two series and compare them.
``` {r exercise6g}
```

## 7) Consider aus_airpassengers, the total number of passengers (in millions) from Australian air carriers for the period 1970-2011.

### a) Use ARIMA() to find an appropriate ARIMA model. What model was selected. Check that the residuals look like white noise. Plot forecasts for the next 10 periods.
``` {r exercise7a}
```

### b) Write the model in terms of the backshift operator.
``` {r exercise7b}
```

### c) Plot forecasts from an ARIMA(0,1,0) model with drift and compare these to part a.
``` {r exercise7c}
```

### d) Plot forecasts from an ARIMA(2,1,2) model with drift and compare these to parts a and c. Remove the constant and see what happens.
``` {r exercise7d}
```

### e) Plot forecasts from an ARIMA(0,2,1) model with a constant. What happens?
``` {r exercise7e}
```

## 8) For the United States GDP series (from global_economy):

### a) if necessary, find a suitable Box-Cox transformation for the data;
``` {r exercise8a}
```

### b) fit a suitable ARIMA model to the transformed data using ARIMA();
``` {r exercise8b}
```

### c) try some other plausible models by experimenting with the orders chosen;
``` {r exercise8c}
```

### d) choose what you think is the best model and check the residual diagnostics;
``` {r exercise8d}
```

### e) produce forecasts of your fitted model. Do the forecasts look reasonable?
``` {r exercise8e}
```

### f) compare the results with what you would obtain using ETS() (with no transformation).
``` {r exercise8f}
```
